Time-series modeling of epidemics in complex populations: detecting changes in incidence volatility over time

### 2025-02-14 ms-3/ms.tex
Update ms.tex

@@

% Metadata
\title{Time-series modeling of epidemics in complex populations: detecting changes in incidence volatility over time}
\author[1]{Rachael Aber}
\author[2]{Yanming Di}
\author[1,3]{Benjamin D. Dalziel}
\affil[1]{Department of Integrative Biology, Oregon State University, Corvallis, Oregon, USA}
\affil[2]{Department of Statistics, Oregon, Oregon State University, Corvallis, Oregon, USA}
\affil[3]{Department of Mathematics, Oregon State University, Corvallis, Oregon, USA}
\date{} % No date for submissions

@@

With both upper and lower $\theta$ thresholds--—corresponding to Poisson-like and zero tolerances, respectively—--maximum likelihood estimates (MLEs) of $\theta$ beyond these thresholds exhibited unbounded behavior. When $\theta$ exceeded the upper threshold, corresponding to processes operationally identical to a Poisson process, the MLE tended to grow arbitrarily large, with the likelihood function reaching its maximum at the upper boundary of the calculated domain. Conversely, when $\theta$ fell below the lower threshold, representing extreme overdispersion with probability mass concentrated near zero, the MLE approached zero, and the likelihood function peaked at the lower boundary of the domain. This behavior reflects the inability of the model to reliably estimate $\theta$ when it lies outside the specified thresholds (Fig.~\ref{fig1}).

@@

Simulations indicate that the LRT framework accurately detects changes in dispersion, with p-values converging to 0.5 as the effect size approaches 0, reflecting the uniform distribution of p-values under the null hypothesis, and decreasing toward zero as the effect size increases (Fig.~\ref{fig1}). The framework is also robust to the range of population sizes present in the empirical data—county populations ranged from approximately 48 thousand to 9.9 million, and we tested the framework on simulated populations between 10 thousand and 10 million. Across this range, the method produced accurate estimates for $\theta$ within $10^{-2} \leq \theta \leq 10^2$, encompassing all operationally relevant values for COVID-19 incidence data and many other infectious diseases. Lower values would concentrate the probability mass function (PMF) for cases almost entirely on 0, while higher values effectively correspond to a Poisson distribution.

@@

Applying the method to COVID-19 cases in US counties enabled investigation of changes in dispersion in relation to both observed trends and model expectations based on case counts (Figure~\ref{fig2}).  
Periods of increased case count variability corresponded with decreases in $\theta$, indicating that dispersion was dynamic. 
Changes in dispersion exhibited both expected and unexpected patterns of variation relative to standard theory.  
In some instances, $\theta$ varied inversely with incidence, consistent with standard epidemic theory, while in other periods, deviations from this expectation occurred, potentially signaling shifts in underlying transmission dynamics.  
Notably, significant changes in the dispersion parameter were observed during major epidemic transitions.
For example, during the beginning of 2022, and at the end of the time series, when the pandemic was transitioning toward endemicity as the landscape of susceptibility was evolving~\citep{lavine2021immunological}.  
The landscape of susceptibility was evolving as a larger proportion of cases involved reinfections.  
These findings underscore the complex behavior of the dispersion parameter, which not only varied with changes in case count regimes but also revealed departures from the model expectations described by Eq.~1, which are consistent with changes in the underlying drivers of transmission.

@@

Dispersion increased markedly around the peaks in incidence during the major 2022 wave, from late December 2021 to early February 2022 (Figure~\ref{fig3}).
This is in strong contrast to standard epidemic theory which predicts that dispersion should decrease as incidence rises.
A high concentration of low $p$-values around peak incidence (Figure~\ref{fig3}) corroborates widespread changes in $\theta$ across counties, reinforcing the statistical significance of this pattern.
While these $p$-values should be corrected for multiple testing if used for inference rather than visualization, the overall trend suggests a systematic departure from theoretical expectations.
Highly overdispersed patterns were also observed more frequently later in the time series, pointing to increasing heterogeneity in transmission, susceptibility, and reporting during the later phases of the pandemic. 
In both the 2022 wave and later in the pandemic, localized surges indicated by higher dispersion may have played a larger role in pandemic dynamics than expected, including potentially placing increased demand for surge capacity in hospitals and testing centers.

@@

>>>>

% Redefine the \figurename to make it bold
\renewcommand{\figurename}{\textbf{Figure}}

@@

% Metadata
\title{Time-series modeling of epidemics in complex populations: detecting changes in incidence volatility over time}
\author[1,2,3]{Rachael Aber}
\author[2]{Yanming Di}
\author[1,4]{Benjamin D. Dalziel}
\affil[1]{Department of Integrative Biology, Oregon State University, Corvallis, Oregon, USA}
\affil[2]{Department of Statistics, Oregon, Oregon State University, Corvallis, Oregon, USA}
\affil[3]{Exponent, Inc., Bellevue, Washington, USA}
\affil[4]{Department of Mathematics, Oregon State University, Corvallis, Oregon, USA}
\date{} % No date for submissions

@@

With both upper and lower $\theta$ thresholds--—corresponding to Poisson-like and zero tolerances, respectively—--maximum likelihood estimates (MLEs) of $\theta$ beyond these thresholds exhibited unbounded behavior. When $\theta$ exceeded the upper threshold, corresponding to processes operationally identical to a Poisson process, the MLE tended to grow arbitrarily large, with the likelihood function reaching its maximum at the upper boundary of the calculated domain. Conversely, when $\theta$ fell below the lower threshold, representing extreme overdispersion with probability mass concentrated near zero, the MLE approached zero, and the likelihood function peaked at the lower boundary of the domain. This behavior reflects the inability of the model to reliably estimate $\theta$ when it lies outside the specified thresholds (Fig~\ref{fig1}).

@@

Simulations indicate that the LRT framework accurately detects changes in dispersion, with p-values converging to 0.5 as the effect size approaches 0, reflecting the uniform distribution of p-values under the null hypothesis, and decreasing toward zero as the effect size increases (Fig~\ref{fig1}). The framework is also robust to the range of population sizes present in the empirical data—county populations ranged from approximately 48 thousand to 9.9 million, and we tested the framework on simulated populations between 10 thousand and 10 million. Across this range, the method produced accurate estimates for $\theta$ within $10^{-2} \leq \theta \leq 10^2$, encompassing all operationally relevant values for COVID-19 incidence data and many other infectious diseases. Lower values would concentrate the probability mass function (PMF) for cases almost entirely on 0, while higher values effectively correspond to a Poisson distribution.

@@

Applying the method to COVID-19 cases in US counties enabled investigation of changes in dispersion in relation to both observed trends and model expectations based on case counts (Fig~\ref{fig2}).  
Periods of increased case count variability corresponded with decreases in $\theta$, indicating that dispersion was dynamic. 
Changes in dispersion exhibited both expected and unexpected patterns of variation relative to standard theory.  
In some instances, $\theta$ varied inversely with incidence, consistent with standard epidemic theory, while in other periods, deviations from this expectation occurred, potentially signaling shifts in underlying transmission dynamics.  
Notably, significant changes in the dispersion parameter were observed during major epidemic transitions.
For example, during the beginning of 2022, and at the end of the time series, when the pandemic was transitioning toward endemicity as the landscape of susceptibility was evolving~\citep{lavine2021immunological}.  
The landscape of susceptibility was evolving as a larger proportion of cases involved reinfections.  
These findings underscore the complex behavior of the dispersion parameter, which not only varied with changes in case count regimes but also revealed departures from the model expectations described by Eq.~1, which are consistent with changes in the underlying drivers of transmission.

@@

Dispersion increased markedly around the peaks in incidence during the major 2022 wave, from late December 2021 to early February 2022 (Fig~\ref{fig3}).
This is in strong contrast to standard epidemic theory which predicts that dispersion should decrease as incidence rises.
A high concentration of low $p$-values around peak incidence (Fig~\ref{fig3}) corroborates widespread changes in $\theta$ across counties, reinforcing the statistical significance of this pattern.
While these $p$-values should be corrected for multiple testing if used for inference rather than visualization, the overall trend suggests a systematic departure from theoretical expectations.
Highly overdispersed patterns were also observed more frequently later in the time series, pointing to increasing heterogeneity in transmission, susceptibility, and reporting during the later phases of the pandemic. 
In both the 2022 wave and later in the pandemic, localized surges indicated by higher dispersion may have played a larger role in pandemic dynamics than expected, including potentially placing increased demand for surge capacity in hospitals and testing centers.

@@

\section*{Supporting Information}

\paragraph*{S1 Text.}
\label{S1_Text}
{\bf Derivation of the relationship between the disperion parameter and the mean crowding parameter.}
